---
title: 'EEMB 179: Week 4: Homework'
author: "Kerri Luttrell"
date: "2/1/2021"
output:
  pdf_document: default
  html_document: default
---
# PART THREE: HOMEWORK
Prep
```{r}
library(deSolve)

```


Second, let's specify the initial conditions for our simulation
```{r}
N <- 100   # population size; because dN/dt = 0, this doesn't change
I0 <- 1		# initial infected individual
S0 <- N - I0	# Assuming this is a new disease, everyone else is susceptible
R0 <- 0 	# Assuming this is a new disease, there are no recovered individuals in the population
```

Third, we'll set up our storage variables. Note: Now we'll need one for each "state" of the population.
```{r}
tset <- seq(from = 0, to = 60, length.out = 20000)# time steps
#set up our holding vectorss for each of our variables : N I S R
N.simu1 <- NaN*tset; N.simu1[1] <- N

I.simu1 <- NaN*tset#infectious
I.simu1[1] <- I0

S.simu1 <- NaN*tset#susceptible
S.simu1[1] <- S0

R.simu1 <- NaN*tset#recovered
R.simu1[1] <- 0

# semi colon separates two lines of code without phsyically separating them
N.simu.a <- NaN*tset; N.simu1[1] <- N
N.simu.b <- NaN*tset; N.simu1[1] <- N
N.simu.c <- NaN*tset; N.simu1[1] <- N

I.simu.a <- NaN*tset#infectious
I.simu.a[1] <- I0
I.simu.b <- NaN*tset#infectious
I.simu.b[1] <- I0
I.simu.c <- NaN*tset#infectious
I.simu.c[1] <- I0

S.simu.a <- NaN*tset#susceptible
S.simu.a[1] <- S0
S.simu.b <- NaN*tset#susceptible
S.simu.b[1] <- S0
S.simu.c <- NaN*tset#susceptible
S.simu.c[1] <- S0

R.simu.a <- NaN*tset#recovered
R.simu.a[1] <- 0
R.simu.b <- NaN*tset#recovered
R.simu.b[1] <- 0
R.simu.c <- NaN*tset#recovered
R.simu.c[1] <- 0
```


1. Make a single Incidence graph (plot number of infected individuals -- $I$ -- over time) that overlays the following scenarios for a population of $N$ = 100. Be sure to include a legend so that we can differentiate between them!
```{r}
###########################################################
# 1a. With a black line: beta = 0.01, gamma = 0.1, p = 0 #
###########################################################
beta <- 0.01   
gamma <- 0.1 
p <- 0

for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	S <- S.simu.a[i-1] 
	I <- I.simu.a[i-1]
	R <- R.simu.a[i-1]
	dS <- (-beta*S*I)*dt
	dI <- (beta*S*I-gamma*I)*dt
	dR <- (gamma*I)*dt
	dN <- dS+dI+dR  
	S.simu.a[i] <- S + dS 
	I.simu.a[i] <- I + dI
	R.simu.a[i] <- R + dR
	N.simu.a[i] <- N + dN
}

##########################################################
# 1b. With a blue line: beta = 0.05, gamma = 0.1, p = 0 #
##########################################################
 beta.b <- 0.05   

 for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	S.b <- S.simu.b[i-1]  
	I.b <- I.simu.b[i-1]
	R.b <- R.simu.b[i-1]
	dS.b <- (-beta.b*S.b*I.b)*dt 
	dI.b <- (beta.b*S.b*I.b-gamma*I.b)*dt
	dR.b <- (gamma*I.b)*dt
	dN.b <- dS.b+dI.b+dR.b  
	S.simu.b[i] <- S.b + dS.b 
	I.simu.b[i] <- I.b + dI.b
	R.simu.b[i] <- R.b + dR.b
	N.simu.b[i] <- N + dN.b
 }	
 
# ###########################################################
# # 1c. With a green line: beta = 0.01, gamma = 0.3, p = 0 #
# ###########################################################
 gamma.c <- 0.3 
   

 for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	S.c <- S.simu.c[i-1]  
	I.c <- I.simu.c[i-1]
	R.c <- R.simu.c[i-1]
	dS.c <- (-beta*S.c*I.c)*dt 
	dI.c <- (beta*S.c*I.c-gamma.c*I.c)*dt
	dR.c <- (gamma.c*I.c)*dt
	dN.c <- dS.c+dI.c+dR.c  
	S.simu.c[i] <- S.c + dS.c 
	I.simu.c[i] <- I.c + dI.c
	R.simu.c[i] <- R.c + dR.c
	N.simu.c[i] <- N + dN.c
 }	

plot(x = tset, y = I.simu.b, type = 'l', las = 1, col = 'blue', xlab = 'Time', ylab='Incidence')
lines(x = tset, y = I.simu.a, col = 'black')
lines(x = tset, y = I.simu.c, col = 'green')
legend(x = max(tset)*0.6, y = max(I.simu.a)*0.9, legend = c('1A Line','1B Line','1C Line'), lwd = 2, col = c('black', 'blue', 'green'))

```

/1 point black line

/1 point blue line

/1 point green line

/1 point legend

/1 point axes labels 

= /5 points total

```{r}
# 2. Describe the differences between the scenarios you plotted above.
#  2a. What is the effect of decreasing infectiousness? 

# By decreasing infectivity of the pathogen and holding all else equal, the incidence of the disease is prolonged. The intial slope is more gradual in plot 1A compared ot plot 1B, which have beta values of 0.1 and 0.5, respectively. Population 1B has a higher maximum for disease incidence because the ratio of beta to gamma is higher, so less individuals are able to move from the infected to the recovered class per new infection.

#  2b. What is the effect of increasing recovery rates? 
# By increasing recovery rates the incidence of the disease is decreased and the disease dies out of the population faster. The recovery outpaces the infection rate in popualtion 1C moreso than in population 1A, so the maximum incidence of the grpah is much less. Also, increasing the line for 1C is always below the line for 1A, which means that adding up the total incidence, less individuals will end up getting infected in 1C overall.


```

/2 points for the effect of decreasing infectiousness

/2 points for the effect of increasing infectiousness

= /4 points total

3. Calculate $R_0$ for each of each of the disease scenarios in Question 1, assuming an initial susceptible population size $S$ = 100. What vaccination proportion ($p$) for each would be required to prevent a disease outbreak?
```{r}
#A
R_0.a <-  beta * N / gamma
p_crit.a <- 1 - 1/R_0.a
p_crit.a
#0.9

#B
R_0.b <-  beta.b * N / gamma
p_crit.b <- 1 - 1/R_0.b
p_crit.b
#0.98

#C
R_0.c <-  beta * N / gamma.c
p_crit.c <- 1 - 1/R_0.c
p_crit.c
#0.7

# Why are we using N in place of S in the above formulation?

# The disease is brand new so no members of the population have immunity to it at the initial timepoint. Thus, the whole population is susceptible and we can use N in place of S.


```

/2 points for work + answer for $p_crit$ scenario 1

/2 points for work + answer for $p_crit$ scenario 2

/2 points for work + answer for $p_crit$ scenario 3

/1 points for $N$ ~ $S$ answer

= /7 points total

4. Run three simulations to check your answers for Part 3. For each, make an incidence diagram (total of 3 incidence diagrams) comparing the unvaccinated scenario (using a black line) with the vaccinated scenario (using a blue line). Don't forget to (1) label your plots so that we know which corresponds to which scenario, and (2) include a legend indicating which line corresponds to vaccinated vs. unvaccinated.
```{r fig.height=10, fig.width=5}
par(mfcol = c(3, 1))
#SCENARIO A
VA <- N*p_crit.a   
S0.v <- N - VA - I0  


N.simu4.a <- NaN*tset; N.simu4.a[1] <- N
S.simu4.a <- NaN*tset; S.simu4.a[1] <- S0.v
I.simu4.a <- NaN*tset; I.simu4.a[1] <- I0
R.simu4.a <- NaN*tset; R.simu4.a[1] <- N - I0 - S0.v - VA

for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	S.4a <- S.simu4.a[i-1]
	I.4a <- I.simu4.a[i-1]
	R.4a <- R.simu4.a[i-1]
	dS.4a <- (-beta*S.4a*I.4a)*dt
	dI.4a <- (beta*S.4a*I.4a-gamma*I.4a)*dt
	dR.4a <- (gamma*I.4a)*dt
	dN.4a <- dS.4a+dI.4a+dR.4a  
	S.simu4.a[i] <- S.4a + dS.4a
	I.simu4.a[i] <- I.4a + dI.4a
	R.simu4.a[i] <- R.4a + dR.4a
	N.simu4.a[i] <- N + dN.4a	
}

plot(x = tset, y = I.simu.a, type = 'l', las = 1, col = 'black', xlab = 'Time', ylab='Incidence', main= 'Scenario A',  ylim= c(0,90))
lines(x = tset, y = I.simu4.a, col = 'blue')
legend(x = max(tset)*0.6, y = 50, legend = c('Unvaccinated','Vaccinated'), lwd = 2, col = c('black', 'blue'))

#SCENARIO B

VB <- N*p_crit.b   
S0.vb <- N - VB - I0  


N.simu4.b <- NaN*tset; N.simu4.b[1] <- N
S.simu4.b <- NaN*tset; S.simu4.b[1] <- S0.vb
I.simu4.b <- NaN*tset; I.simu4.b[1] <- I0
R.simu4.b <- NaN*tset; R.simu4.b[1] <- N - I0 - S0.vb - VA

for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	S.4b <- S.simu4.b[i-1]
	I.4b <- I.simu4.b[i-1]
	R.4b <- R.simu4.b[i-1]
	dS.4b <- (-beta.b*S.4b*I.4b)*dt
	dI.4b <- (beta.b*S.4b*I.4b-gamma*I.4b)*dt
	dR.4b <- (gamma*I.4b)*dt
	dN.4b <- dS.4b+dI.4b+dR.4b  
	S.simu4.b[i] <- S.4b + dS.4b
	I.simu4.b[i] <- I.4b + dI.4b
	R.simu4.b[i] <- R.4b + dR.4b
	N.simu4.b[i] <- N + dN.4b	
}

plot(x = tset, y = I.simu.b, type = 'l', las = 1, col = 'black', xlab = 'Time', ylab='Incidence', main= 'Scenario B', ylim= c(0,90))
lines(x = tset, y = I.simu4.b, col = 'blue')
legend(x = max(tset)*0.6, y = 50, legend = c('Unvaccinated','Vaccinated'), lwd = 2, col = c('black', 'blue'))
?ylim
#SCENARIO C

VC <- N*p_crit.c   
S0.vc <- N - VC - I0  


N.simu4.c <- NaN*tset; N.simu4.c[1] <- N
S.simu4.c <- NaN*tset; S.simu4.c[1] <- S0.vb
I.simu4.c <- NaN*tset; I.simu4.c[1] <- I0
R.simu4.c <- NaN*tset; R.simu4.c[1] <- N - I0 - S0.vb - VA

for(i in 2:length(tset)){
	dt <- tset[i]-tset[i-1]
	S.4c <- S.simu4.c[i-1]
	I.4c <- I.simu4.c[i-1]
	R.4c <- R.simu4.c[i-1]
	dS.4c <- (-beta*S.4c*I.4c)*dt
	dI.4c <- (beta*S.4c*I.4c-gamma.c*I.4c)*dt
	dR.4c <- (gamma.c*I.4c)*dt
	dN.4c <- dS.4c+dI.4c+dR.4c  
	S.simu4.c[i] <- S.4c + dS.4c
	I.simu4.c[i] <- I.4c + dI.4c
	R.simu4.c[i] <- R.4c + dR.4c
	N.simu4.c[i] <- N + dN.4c	
}

plot(x = tset, y = I.simu.c, type = 'l', las = 1, col = 'black', xlab = 'Time', ylab='Incidence', main= 'Scenario C', ylim= c(0,90))
lines(x = tset, y = I.simu4.c, col = 'blue')
legend(x = max(tset)*0.5, y = 30, legend = c('Unvaccinated','Vaccinated'), lwd = 2, col = c('black', 'blue'))
```


```

/3 points for unvaccinated lines

/3 points for vaccinated lines

/3 points for legend

/3 points for axes 

= /12 points total (3 plots)


 5. Explain how R_0 and p depend upon:
 R_0 =  beta * (N / gamma)
 p_crit = 1- 1/R_0
 
  5a. infectiousness, 

When infectiousness increases, RO increases asymptotically.
When infectiousness increases, p_crit increases asymptotically.

  5b. recovery time, and 

When recovery time increases, RO decreases linearly.
When recovery time increases, p_crit decreases linearly.

  5c. population size (assuming N = S, when the disease has not yet arrived). 

When population size increseases, RO increases asymptotically.
When population size increseases, p_crit increases asymptotically.

 For each of these (for a total of 6 plots -- don't forget to label the axes of each one!): 
	- create a plot of R_0 (y-axis) vs. the parameter/variable of interest (x-axis),
  - create a plot of p (y-axis) vs. the parameter of interest (x-axis), and
	- explain the shapes of the graphs based on the biology of disease spread. Do you notice anything odd about the answer to 5c?


```{r fig.height=10, fig.width=7}
beta_set <- seq(from = 0.001, to = 0.1, length.out = 100)
gamma_set <- seq(from = 0.001, to = 0.1, length.out = 100)
R0_set <- beta_set*N/gamma
Nset <- seq(from = 0, to = 100, length.out = 100)


par( mfrow= c(3,1))

#R_O x Beta
plot(x = beta_set, y = R0_set, type = 'l', xlab = 'Infectiousness (beta)', ylab = 'Basic Reproductive Number (R_0)', las = 1, lwd = 2)

R0g_set <- beta*N/gamma_set
# R_O x Gamma
plot(x = gamma_set, y = R0g_set, type = 'l', xlab = 'Recovery Time (gamma)', ylab = 'Basic Reproductive Number (R_0)', las = 1, lwd = 2)

R0n_set <- beta*Nset/gamma
# R_O x Population Size
plot(x = Nset, y = R0n_set, type = 'l', xlab = 'Population Size (N)', ylab = 'Basic Reproductive Number (R_0)', las = 1, lwd = 2)
```

```{r fig.height=10, fig.width=7}
par( mfrow= c(3,1))
p_crit_set <- 1 - 1/R0_set

# p_crit x Beta
plot(x = beta_set, y = p_crit_set, type = 'l', xlab = 'Infectiousness (beta)', ylab = 'Critical Vaccination Rate (p_crit)', las = 1, lwd = 2)

p_critg_set <- 1 - 1/R0g_set

# p_crit x Gamma
plot(x = gamma_set, y = p_critg_set, type = 'l', xlab = 'Recovery Time (gamma)', ylab = 'Critical Vaccination Rate (p_crit)', las = 1, lwd = 2)

p_critn_set <- 1 - 1/R0n_set
# p_crit x Pop Size
plot(x = Nset, y = p_critn_set, type = 'l', xlab = 'Population Size (N)', ylab = 'Critical Vaccination Rate (p_crit)', las = 1, lwd = 2)
```

The pcrit values are negative for p_crit x Pop Size because at a low population size (N<10) with a beta of 0.01 and gamma of 0.1 the R0 value is less than 1. So, pcrit = 1 - 1/ RO would be a negative number. 

/6 points for explanation of how R_0 and p depend upon infectiousness, recovery time, and population size.

/6 point per simulation line 

/6 point per set of axes 

/2 point for explanation of negative p_crit values in answer to 5c

= /20 points total (6 plots)
